---
title: "DNA-seq data analysis"
author: "Givanna Putri"
date: "2023-10-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Analysis for the DNA-seq 8k and 10k library of MCF7 cell line tagged with ClonMapper protocol.

```{r}
library(CloneDetective)
library(data.table)
library(ggplot2)
library(scales)
```

## Read in data

NextClone and pycashier output.

The command used to generate NextClone output is available in `code` folder.

```{r}
clones_nxclone <- fread("data/nextclone_out/dnaseq_clone_barcode_counts_20231114.csv")
# The samples in the sample_name column is way too complicated. 
# Let's create a new column.
clones_nxclone[, sample_name_simple := gsub("vexGFP-", "", gsub("_.*", "", sample_name))]
clones_nxclone[, sample_name_simple := factor(sample_name_simple, levels = c("8k", "10k"))]

clones_pycashier <- lapply(c("8k", "10k"), function(samp) {
  # read_count as the count column so we can use count_retained_clones
  dt <- fread(
    file = paste0("data/pycashier_out/", samp, ".tsv"),
    header = FALSE,
    col.names = c("clone_barcode", "read_count")
  )
  dt[, sample := samp]
  return(dt)
})

clones_pycashier <- rbindlist(clones_pycashier)
clones_pycashier[, sample := factor(sample, levels = c("8k", "10k"))]
```

## Number of unique barcodes

Count the number of unique barcodes with at least x number of cells.

```{r}

thresholds <- c(1, 20, 200, 1000)

n_barcodes_nxclone <- count_retained_clones(
    count_data = clones_nxclone,
    thresholds = thresholds,
    grouping_col = "sample_name_simple",
    count_column = "read_count"
)
n_barcodes_nxclone[, tool := 'NextClone']
setnames(n_barcodes_nxclone, "sample_name_simple", "sample")

n_barcodes_pycashier <- count_retained_clones(
    count_data = clones_pycashier,
    thresholds = thresholds,
    grouping_col = "sample",
    count_column = "read_count"
)
n_barcodes_pycashier[, tool := 'PyCashier']

n_barcodes <- rbind(n_barcodes_nxclone, n_barcodes_pycashier)
```

Plot bar chart that compares the number of unique barcodes retrieved by both
workflows.

```{r fig.width=10, fig.height=8}
n_barcodes_long <- melt(n_barcodes, id.vars = c("sample", "tool"), 
                        variable.name = "filtering_threshold",
                        value.name = "n_barcode")

filtering_threshold_levels <- paste(">=", thresholds, "cells")
n_barcodes_long[, filtering_threshold := factor(
  gsub("_"," ",gsub("at_least_", ">= ", filtering_threshold)),
  levels = filtering_threshold_levels
)]

ggplot(n_barcodes_long, aes(x=factor(filtering_threshold), y=n_barcode, fill=tool)) +
  geom_bar(stat="identity", position=position_dodge()) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
    legend.position="bottom"
  ) +
  facet_wrap(~ sample) +
  scale_y_continuous(breaks = pretty_breaks(n=10), label = label_comma(accuracy = 1)) +
  labs(
    y = "No. of barcodes",
    x = "Filtering thresholds",
    fill = "Pipeline",
    title = "No. of barcodes retrieved for DNA-seq data"
  )
```

## Ordered abundance plot

To show the proportion of barcode's frequency.

```{r}
clones_nxclone_filtered <- remove_clones_below_threshold(
    count_data = clones_nxclone,
    threshold = 20,
    count_column = "read_count"
)
clones_nxclone_filtered <- convert_count_to_proportion(
  count_data = clones_nxclone_filtered,
  grouping_col = "sample_name_simple",
  count_column = "read_count"
)
```


```{r}
plt <- draw_ordered_abundance_plot(
  count_data = clones_nxclone_filtered, 
  facet_column = "sample_name_simple",
  y_axis_column = "read_proportion"
)
plt <- plt +
  geom_point(size=0.5, colour='red') +
  labs(
    title = "Proportion of cells assigned to each barcode",
    subtitle = "Barcode IDs are numerically assigned in order of read proportion",
    x = "Numerical barcode ID",
    y = "Proportion of cells per sample"
  )
plt
```

## Using NGS data to plan single cell experiment

Let's say we want to sequence 10,000 cells. 
Based on our NGS data, can we predict what will happen to our clone barcodes?
Will we get enough representations?

Do projection by calculating proportion and multiply by amount of cells to be projected to.

How many cells we will get per clone?

```{r}
clones_nxclone_proportion <- project_clones(
  count_data = clones_nxclone,
  grouping_col = "sample_name_simple",
  count_column = "read_count",
  project_amnt = c(10000, 20000)
)

plt <- draw_ordered_abundance_plot(
  count_data = clones_nxclone_proportion,
  y_axis_column = 'projected_20000_confidence_1',
  facet_column = 'sample_name_simple'
) + 
  geom_point(size = 0.5, colour='forestgreen') +
  labs(
    y = 'Number of cells',
    title = 'Number of cells assigned to each barcode',
    subtitle = 'Cell counts computed after projection to 20,000 cells'
  )
plt
```

```{r}
plt <- draw_ordered_abundance_plot(
  count_data = clones_nxclone_proportion,
  y_axis_column = 'projected_10000_confidence_1',
  facet_column = 'sample_name_simple'
) + 
  geom_point(size = 0.5, colour='blue') +
  labs(
    y = 'Number of cells',
    title = 'Number of cells assigned to each barcode',
    subtitle = 'Cell counts computed after projection to 10,000 cells'
  )
plt
```

How many clones that contain at least 10, 20, 50, 100 cells?

```{r fig.width=10, fig.height=8}
thresholds <- seq(10, 30, 5)

proj_n_clones_retained <- count_retained_clones(
  count_data = clones_nxclone_proportion,
  thresholds = thresholds,
  grouping_col = "sample_name_simple",
  count_column = "projected_10000_confidence_1"
)
names(proj_n_clones_retained) <- c("sample", paste(">=", thresholds, "cells"))

proj_n_clones_retained_long <- melt(proj_n_clones_retained, id.vars = c("sample"), 
                        variable.name = "filtering_threshold",
                        value.name = "n_barcode")


ggplot(proj_n_clones_retained_long, aes(x=factor(filtering_threshold), y=n_barcode)) +
  geom_bar(stat="identity", position=position_dodge(), fill="darkorange") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
    legend.position="bottom"
  ) +
  facet_wrap(~ sample) +
  scale_y_continuous(breaks = pretty_breaks(n=6),label = label_comma(accuracy = 1)) +
  labs(
    y = "No. of barcodes",
    x = "Filtering thresholds",
    fill = "Pipeline",
    title = "No. of barcodes estimated for DNA-seq data",
    subtitle = "When capturing 10,000 cells"
  )

```


What are the frequency of top 200 clone barcodes?
We can present this by computing the number of cells tagged by top 200 clone barcodes.

```{r}
top_threshold <- 200
top_barcodes <- get_top_barcodes_and_cum_sum(
  count_data = clones_nxclone_proportion,
  count_column = "projected_10000_confidence_1",
  grouping_col = "sample_name_simple",
  top_threshold = top_threshold
)
```

Create a line chart that show cumulative number of cells.

```{r fig.width=10, fig.height=10}
ggplot(top_barcodes, aes(x=barcode_rank, y=cum_sum_projected_10000_confidence_1, 
                         group=sample_name_simple, colour = sample_name_simple)) +
  geom_line(linewidth=1) +
  theme_bw(base_size = 16) +
  scale_y_continuous(breaks = pretty_breaks(n=10), labels = label_comma(accuracy = 1)) +
  scale_x_continuous(breaks = pretty_breaks(n=10)) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    legend.position = "bottom"
  ) +
  labs(
    y = "Number of cells",
    x = "Barcode Ranking",
    title = paste("Projected no. of cells for top", top_threshold, "most abundant clones"),
    subtitle = "When capturing 10,000 cells",
    colour = "Library"
  )
```



