---
title: "sc_data_analysis_withCloneDetective"
author: "Givanna Putri"
date: "2023-11-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Working directory for NextClone run: `/vast/projects/Goel_senescence/nextclone_dev/07_analysis/pilot_dataset/02_run_nextclone/output_20231108_v2`


```{r message=FALSE, warning=FALSE}
library(data.table)
library(DropletUtils)
library(CloneDetective)
library(scater)
```

Load data

```{r}
raw_clone_data <- fread("data/nextclone_out/sc_clone_barcodes_20231109.csv")
```

Convert to expression
```{r}
cell_by_clone_mat <- generate_cell_clone_barcode_matrix(cell_clone_bcode_dt = raw_clone_data)
```

Incorporate the cell by gene matrix.

```{r}
sce <- read10xCounts("data/cellranger_out/filtered_feature_bc_matrix")
sce
```

Compute few simple metrics like average library size per cell.
Compute total transcript molecules detected per cell then compute average
based on the number of cells detected.

```{r}
cell_qc_metrics <- perCellQCMetrics(sce)
summary(cell_qc_metrics$sum)
summary(cell_qc_metrics$detected)
```


Get the 10x cell barcode.

```{r}
valid_cells_10x <- colData(sce)$Barcode
```

Tree map to show the proportion of 10x cells that have 0, 1, 2, .. clones found.

```{r}
plt <- draw_treemap(
    cell_by_clone_matrix = cell_by_clone_mat,
    valid_cells_bcodes = valid_cells_10x
)
plt
```

# Assign clone barcodes to cells

```{r}
sce_with_clone <- assign_and_embed_clones(
    cell_by_gene_mat = sce,
    cell_clone_reads_dt = raw_clone_data,
)
colData(sce_with_clone)

```

Count how many cells assigned to most dominant clones.

```{r}
clone_bcode_criteria <- as.data.table(colData(sce_with_clone))
clone_bcode_criteria <- data.table(table(clone_bcode_criteria$clone_barcode_criteria))

# proportion. of cells with multiple clone barcodes?
clone_bcode_criteria[, multiclone := ! V1 %in% c("single_clone", "no_clones_found")]
clone_bcode_criteria[, .(prop = sum(N) / sum(clone_bcode_criteria$N)), by = multiclone]

# proportion of multiclones assigned to most dominant clone?
clone_bcode_criteria[, .(prop = N / sum(N)), by = multiclone]
```

Export the clone assignments as data.table which can later be saved.

```{r}
clone_assignments <- assign_and_embed_clones(
    cell_by_gene_mat = sce,
    cell_clone_reads_dt = raw_clone_data,
    embed_to_mat = FALSE
)
head(clone_assignments)
```

